---
title: "HW 3"
author: "Student Name"
date: "9/24/2024"
output: 
  html_document:
    number_sections: true
---

#

Let $E[X] = \mu$.  Show that $Var[X] := E[(X-E[X])^2] = E[X^2]-(E[X])^2$.  Note, all you have to do is show the second equality (the first is our definition from class). 

 $E[(X-E[X])^2] = E[(X^2-2XE[X]+E[X]^2)] = E[X^2] -2E[(XE[X])] +(E[X])^2 = E[X^2] -2E[X]^2 + (E[X])^2 = E[X^2]-(E[X])^2$

# 

In the computational section of this homework, we will discuss support vector machines and tree-based methods.  I will begin by simulating some data for you to use with SVM. 

```{r}
library(e1071)
set.seed(1) 
x=matrix(rnorm(200*2),ncol=2)
x[1:100,]=x[1:100,]+2
x[101:150,]=x[101:150,]-2
y=c(rep(1,150),rep(2,50))
dat=data.frame(x=x,y=as.factor(y))
plot(x, col=y)

```


##

Quite clearly, the above data is not linearly separable.  Create a training-testing partition with 100 random observations in the training partition.  Fit an svm on this training data using the radial kernel, and tuning parameters $\gamma=1$, cost $=1$.  Plot the svm on the training data.  

```{r}
set.seed(1)

train_indices <- sample(1:200, 100)
train_dat <- dat[train_indices, ]
test_dat <- dat[-train_indices, ]

svmfit = svm(y~., data=train_dat, kernel = 'radial', cost=1, gamma=1, scale= FALSE)
print(svmfit)
plot(svmfit, train_dat)

```

##

Notice that the above decision boundary is decidedly non-linear.  It seems to perform reasonably well, but there are indeed some misclassifications.  Let's see if increasing the cost ^[Remember this is a parameter that decides how smooth your decision boundary should be] helps our classification error rate.  Refit the svm with the radial kernel, $\gamma=1$, and a cost of 10000.  Plot this svm on the training data. 

```{r}
svmfit = svm(y~., data=train_dat, kernel = 'radial', cost=10000, gamma=1, scale= FALSE)
print(svmfit)
plot(svmfit, train_dat)

```

##

It would appear that we are better capturing the training data, but comment on the dangers (if any exist), of such a model. 

Although this model captures the training data better, it involves a significant increase in the cost function. Increased cost in this case results in a less smooth decision boundary. As a result, the model is prone to overfitting which leads to poor generalization. Thus, the model may behave poorly with new data.

##

Create a confusion matrix by using this svm to predict on the current testing partition.  Comment on the confusion matrix.  Is there any disparity in our classification results?    

```{r}
#remove eval = FALSE in above
table(true=dat[-train_indices,"y"], pred=predict(svmfit, newdata=dat[-train_indices,]))
```
The confusion matrix indicates that 17 points were mis-classified as 2 when their true value is 1 and 3 points were mis-cassified as 1 when their true value is 2. This resuls in a misclassification error rate of (17+3)/100 = 0.2. Therefore, the confusion matrix demonstrates high mis-classification indicating some disparity.

##

Is this disparity because of imbalance in the training/testing partition?  Find the proportion of class `2` in your training partition and see if it is broadly representative of the underlying 25\% of class 2 in the data as a whole.  

```{r}
summary(train_dat)
summary(dat)
```

The training set has a higher proportion of class 2 (29%) compared to the overall dataset (25%). While the 4% difference is small, it may still affect the classification results, especially if the model is sensitive to class imbalances.

##

Let's try and balance the above to solutions via cross-validation.  Using the `tune` function, pass in the training data, and a list of the following cost and $\gamma$ values: {0.1, 1, 10, 100, 1000} and {0.5, 1,2,3,4}.  Save the output of this function in a variable called `tune.out`.  

```{r}

set.seed(1)
tune.out <- tune(svm, 
                 y ~ ., 
                 data = train_dat, 
                 kernel = "radial", 
                 ranges = list(cost = c(0.1, 1, 10, 100, 1000), gamma = c(0.5, 1, 2, 3, 4)))


```

I will take `tune.out` and use the best model according to error rate to test on our data.  I will report a confusion matrix corresponding to the 100 predictions.  


```{r}
table(true=dat[-train_indices,"y"], pred=predict(tune.out$best.model, newdata=dat[-train_indices,]))
```

##

Comment on the confusion matrix.  How have we improved upon the model in question 2 and what qualifications are still necessary for this improved model.  

The new confusion matrix indicates a mis-classification rate of 8/100 = 0.08, a significant decrease from 0.2. However, it is important to ensure the model still generalizes well to other data. It is possible the improved model remains prone to overfitting. This is especially true in this case as we are working with a non-linear model

# 
Let's turn now to decision trees.  

```{r}

library(kmed)
data(heart)
library(tree)

```

## 

The response variable is currently a categorical variable with four levels.  Convert heart disease into binary categorical variable.  Then, ensure that it is properly stored as a factor. 

```{r}
hist(heart$class) #class represents response variable where a value not equal to 0 represents presence of heart diesease

HD = ifelse(heart$class < 1, "No", "Yes")
HD_fac <- as.factor(HD)
heart <- data.frame(heart, HD_fac)
str(heart)

```

## 

Train a classification tree on a 240 observation training subset (using the seed I have set for you).  Plot the tree.  

```{r}
set.seed(101)
train = sample(1:nrow(heart), 240)

tree.heart = tree(HD_fac~.-class, heart, subset=train)
plot(tree.heart)
text(tree.heart, pretty = 0)
```


## 

Use the trained model to classify the remaining testing points.  Create a confusion matrix to evaluate performance.  Report the classification error rate.  

```{r}
tree.pred = predict(tree.heart, heart[-train,], type="class")
with(heart[-train,], table(tree.pred, HD_fac))

(3+8)/(28+3+8+18) #CLASSIFICATION RATE
```

##  

Above we have a fully grown (bushy) tree.  Now, cross validate it using the `cv.tree` command.  Specify cross validation to be done according to the misclassification rate.  Choose an ideal number of splits, and plot this tree.  Finally, use this pruned tree to test on the testing set.  Report a confusion matrix and the misclassification rate.  

```{r}
set.seed(101)
cv.heart = cv.tree(tree.heart, FUN=prune.misclass)
cv.heart

plot(cv.heart$size, cv.heart$dev, type = "b")

prune.heart = prune.misclass(tree.heart, best = 3)
plot(prune.heart)
text(prune.heart, pretty=0)

tree.pred2 = predict(prune.heart, heart[-train,], type="class")
with(heart[-train,], table(tree.pred2, HD_fac))

(4+10)/(4+10+26+17)
```


##

Discuss the trade-off in accuracy and interpretability in pruning the above tree. 

Pruning the tree results in a small increase of about 0.05 in misclassification error. Although accuracy has decreased, we have improved interpretability and generalizability. In this case, the cross-validation minimum occurs at around 3 splits. We have significantly decreased the number of splits compared to the full model, which decreases the risk of overfitting.

## 

Discuss the ways a decision tree could manifest algorithmic bias.  

A decision tree can manifest algorithmic bias when trained on non-representative data. It can also occur if overgrown trees contain splits based on overfitting to class lables like race or sex, which can lead to biased predictions.